Practical Optional Types for Clojure

### 2015-02-21 intro.tex

Programming is an error prone activity, especially in the presence of
multiple languages working together. Over the past decade, it has
become increasingly frequent to extend existing dynamically typed
languages with optional type systems.  The guarantees that a sound
type system provides can be the difference between choosing a language
that supports a type system or one that does not.

% presenting clojure

Clojure is a dynamically typed language built to run on the Java Virtual Machine (JVM)
which has recently gained popularity as an alternative JVM language.
It offers the flexibility of a Lisp dialect, including Common Lisp-style macros,
emphasising a functional style via a standard library of immutable datastructures. 
Clojure has good interoperability with existing Java code, allowing programmers to
use existing Java libaries without leaving Clojure.

%problem?

Clojure

Typed Clojure is a gradual type system for Clojure, based on the
static portion of Typed Racket.

\begin{minted}[]{clojure}
(ann parent [(U nil File) -> (U nil File)])
(defn parent [^File f]
  (when f
    (let [s (.getParent f)]
      (assert s "Expected parent string, got nil")
      (new File s))))
\end{minted}

@@

\begin{minted}[]{clojure}
(defalias Expr
  (U '{:op :if :test Expr :then Expr :else Expr}
     '{:op :const :val Any} ...))
(ann tree-str [Expr -> String])
(defmulti tree-str (fn [e :- Expr] (get :op e)))
(defmethod tree-str :if [{:keys [test then else]}] 
  (str (tree-str test) (tree-str then) (tree-str else)))
(defmethod tree-str :const [{:keys [val]}] 
  (str val))
\end{minted}

\subsection{Contributions}

%The main idea is....... SPJ
In this paper we present a practical type system for Clojure.

\begin{itemize}

  \item We formalise Typed Clojure and prove that typed code
    \begin{itemize}
      \item disallows null-pointer exceptions, (Section ?)
      \item statically resolves Java interoperability (Section ?)
    \end{itemize}
  \item We reveal the surprising synthesis between Clojure multimethods
        and occurrence typing.
  \item We 
\end{itemize}

>>>>

We synthesize these lessons to present \emph{Typed Clojure}, an
optional type system for Clojure. Typed Clojure builds on the core
type checking approach of Typed Racket, an existing gradual type
system for Racket. However, Typed Clojure extends this basic framework
in multiple ways to accommodate the unique idioms and features of
Clojure, producing an expressive synthesis of ideas and demonstrating
a surprising coincidence between the multiple dispatch approach of
Clojure and the occurrence typing framework of Typed Racket.

The essence of Typed Clojure, of course, is Clojure, a dynamically
typed language in the Lisp family built to run on the Java Virtual
Machine (JVM) which has recently gained popularity as an alternative
JVM language.  It offers the flexibility of a Lisp dialect, including
macros, emphasizes a functional style via a
standard library of immutable data structures, and provides
interoperability with existing Java code, allowing programmers to use
existing Java libraries without leaving Clojure.
%
Since its initial release in 2007, Clojure has been widely adopted for
``backend'' development in places where its support for paralleism,
functional programming, and Lisp-influenced abstraction is desired on
the JVM. As a result, it now has an extensive base of existing untyped
programs, many of whose developers can benefit from Typed Clojure. As
a result, Typed Clojure is now used in industry, experience we discuss
in this paper.

\begin{figure}
\begin{minted}[]{clojure}
(ann parent [(U nil File) -> (U nil File)])
(defn parent [^File f]
  (when f
    (let [s (.getParent f)]
      (assert s "Expected parent string, got nil")
      (new File s))))
\end{minted}
\caption{A simple Typed Clojure program}
\label{fig:ex1}
\end{figure}
Figure~\ref{fig:ex1} presents a simple program demonstrating several
aspects of our system, from simple type annotations to explicit
handling of Java's \texttt{null} in interoperation, as well as an
extended form of occurrence typing and Clojure's \emph{type hints},
which are central to Typed Clojure's approach to interoperability.

In the remainder of this paper, we describe Typed Clojure's central
innovations, including Java interoperability, multimethods, and
heterogenously-typed immutable maps, following an example-driven
presentation in Section~\ref{sec:overview}. We then incrementatlly
present a core calculus for Typed Clojure covering all of these
features together in Section~\ref{sec:formal} and prove type
soundness(section \ref{sec:sound}). We then discuss the full
implementation of Typed Clojure, dubbed \texttt{core.typed}, which
extends the formal model in many ways, and the experience gained from
its use in Section~\ref{sec:exp}. Finally, we discuss related work and
conclude.

@@

%% \begin{minted}[]{clojure}
%% (defalias Expr
%%   (U '{:op :if :test Expr :then Expr :else Expr}
%%      '{:op :const :val Any} ...))
%% (ann tree-str [Expr -> String])
%% (defmulti tree-str (fn [e :- Expr] (get :op e)))
%% (defmethod tree-str :if [{:keys [test then else]}] 
%%   (str (tree-str test) (tree-str then) (tree-str else)))
%% (defmethod tree-str :const [{:keys [val]}] 
%%   (str val))
%% \end{minted}

%% \subsection{Contributions}

%% %The main idea is....... SPJ
%% In this paper we present a practical type system for Clojure.

%% \begin{itemize}

%%   \item We formalise Typed Clojure and prove that typed code
%%     \begin{itemize}
%%       \item disallows null-pointer exceptions, (Section ?)
%%       \item statically resolves Java interoperability (Section ?)
%%     \end{itemize}
%%   \item We reveal the surprising synthesis between Clojure multimethods
%%         and occurrence typing.
%%   \item We 
%% \end{itemize}
